---
title: "Intro to Shiny"
author: "Thomas Wood"
date: "04/02/2022"
theme: readable 
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(slickR)
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy()
```

## Create a new shiny app

Open RStudio and go to:

`File > New Project... > New Directory > Shiny Web Application > Enter a directory name and location > Create project` 

R will open a new project for the shiny app, which will contain one file (`app.R`), which will have an example shiny app that you can run by clicking the `Run App` button (top right of the script panel).

This is the home of your app, and any data, pictures, etc. you need for your app to run need to be stored in the same directory so that they can be accessed by the app once it is deployed.

> Note: Make sure you have shiny installed. 

```{r eval=FALSE, echo=TRUE} 
install.packages("shiny")
```

## Clone the workshop

[You can download a copy of this workshop, and all the code here. ](https://github.com/woodtho/CanDev_2022/archive/refs/heads/main.zip) After downloading the file, you can unzip it and then launch the project file (`CanDev_2022.Rproj`) in RStudio.

## App structure

`app.R` has three components:

- a user interface object
- a server function
- a call to the `shinyApp()` function

You can include other code inside your app (e.g., reading data, loading packages, creating static variables). This code is run once (when the app is started), and never again.

The user interface (`ui`) object controls the layout and appearance of your app. The `server` function contains the instructions that a computer needs to build the app. Finally, the `shinyApp` function creates the Shiny app from an explicit UI/server pair.

Here is a high level overview of a shiny app (we will cover each part in more specifics later on):

```{r fig.height=8}
slickR(
  list.files("img", full.names = TRUE, pattern = 'png'),
  height = 700,
  width = '100%'
) %synch%
  (slickR(
    c("RStudio provides this app every time you create a new shiny application.",
      
      "To create the app, we will first need to create the UI object, which contains the HTML that will be displayed on the webpage. You can see what this HTML looks like by printing out the UI object. The HTML in the UI is static, so it is the same for everyone who views the app.",
      
      "The next step is to create the server function, which tracks user inputs and updates the webpage based on changes made by the user. Users will see the same UI, but they will each have their own instance of the server function. This ensures that any changes one user makes to the application will only affect the webpage they are viewing.",
      
      "ShinyApp() combines the UI object with the server function into an app. After running this call, your R session becomes a server and starts hosting your app on your local network. To be able to submit new code, you will need to stop hosting the webpage by pressing the stop button in the console, just closing the browser isn't enough.",
      
      "Besides the three required components, you can add other R code that will be executed once when starting the app. For example, reading a table from a file, creating static variables, or loading required packages. In this case we are just loading the shiny library.",
      
      "Here is the example app in action.",
      
      "We have one slider input and one plot output in the UI. It is important to note that the plotOutput() function does not actually create the plot. Instead, it just creates a space in your UI for the plot. The plot is created by the server.",
      
      "You can read values from inputs using the input object and write new html to be outputs using the output object from within the server.",
      
      "Here is an application in a steady state, waiting for the user to interact with it.",
      
      "When we change the input, we update the value stored in the input object.",
      
      "The server constantly monitors the inputs and notices that one has changed.",
      
      "The server will re-run any code that references the value from the updated input. In this case, the plot uses the bins input to determine how many bins to use for the histogram, so the server re-runs the plot code.",
      
      "As soon as the server finishes updating the plot and rendering it into HTML, the results are saved into the output object.",
      
      "When the new HTML is stored in the output object, the webpage is refreshed and the outputs are updated. Once this process is complete, the app enters a steady state and awaits a new input."
    ),
    slideType = 'p'
  ) + settings(arrows = FALSE))


```



## Reactivity

Shiny is about making it easy to wire up input values from a web page, making them available to you in R, and have the results of your R code be written as output values back out to the web page.

The reactivity framework used by shiny means that:

- You can change inputs at any time, and the outputs that are impacted will updated immediately to reflect those changes.
- Shiny tracks "dependencies" between inputs and outputs automatically, so you don't have to write code to track changes.
- Using a list-like syntax, you can access inputs from the browser in the server, and write outputs from the server back to the webpage

## App template

> You can copy and paste almost all of the code in this workshop into your R console to launch a shiny app. Please ensure that you have installed the required packages.

```{r eval=FALSE, echo=TRUE}
install.packages("shiny")
```

- Whenever you create a shiny app, you should work from a template. 
- RStudio opens all new shiny apps with a simple app that can then be modified.

Here is the basic skeleton of a shiny app:

```{r eval=FALSE, echo=TRUE}
library(shiny)

# Any code that should be executed once when the app starts. E.g., reading in a
# data file, creating static environment variables.

# Code that creates the UI of the app
ui <- fluidPage(
  # Title for your app
  titlePanel(), 
  
  sidebarLayout(
    
    # Generally, inputs are put into the sidebar
    sidebarPanel(
      
      # Inputs go here
      
      ),
    
    # And outputs are put into the mainPanel
    mainPanel(
      
      # Outputs go here
      
      )
    )
  )

# The server takes inputs from the webpage, updates outputs, and send the
# outputs back to the webpage
server <- function(input, output) { 
  
  # Server code goes here
  
  }

# Used to launch the app after defining the UI and server.
shinyApp(ui, server)

```

There are other templates you can use as well; but a fluid page with a sidebar layout is a good option. Fluid pages will reflow content in case there isn't enough horizontal space. The sidebar layouts separates the page into two panels (sidebar and main), making it easier to organize inputs and outputs.

With functions like `fluidRow()` and `column()` you can have more control over the layout of an app, but for the moment we will just arrange the app into the sidebar and main panel.

## Static UI elements

Aside from reactive elements, you might also want to include static elements in your UI. Static elements are things like text, or images that do not change with user input. Common tags are available from the shiny package (`div()`, `h1()`, `a()`, `br()`). Less common tags can be inserted with the `tags` object (e.g., `tags$abbr()`, `tags$figcaption()`). 

Here is an example app with just static content:

```{r eval=FALSE, echo=TRUE}
library(shiny)

# Any code that should be executed once when the app starts. E.g., reading in a
# data file, creating static environment variables.

# Code that creates the UI of the app
ui <- fluidPage(# Title for your app
  titlePanel("A first level heading title"),
  
  sidebarLayout(
    # Generally, inputs are put into the sidebar
    sidebarPanel(
      h2("A second level heading"),
      "This small Shiny application demonstrates some static UI elements."
    ),
    
    # And outputs are put into the mainPanel
    mainPanel(
      h2("A second level header in the main panel"),
      p('My first paragraph, with some ', strong('bold'), ' text.')
      )
    )
  )

# The server takes inputs from the webpage, updates outputs, and send the
# outputs back to the webpage
server <- function(input, output) { 
  
  # Because the app is static, there is no server code!
  
  }

# Used to launch the app after defining the UI and server.
shinyApp(ui, server)

```

- Notice that the elements we added inside the sidebarPanel and mainPanel are separated with commas!
- [See this tutorial for more examples of adding static content to your interface, including adding images and aligning text.](http://shiny.rstudio-staging.com/tutorial/written-tutorial/lesson2/) 

## Building your app around **Inputs and outputs**

Until now, the apps have been simple and contained mostly static elements that we created in the UI. Shiny, however, is all about making things interactive. We do this by using inputs and outputs.

### Inputs

- Are created with an `*Input()` function
- They create HTML for your app's UI and send values from the web page to the server

| Function             | Collects                                                                                                 |
|----------------------|----------------------------------------------------------------------------------------------------------|
| `actionButton()`       | A button press                                                                                           |
| `checkboxInput()`      | The state of a check box (TRUE/FALSE)                                                                     |
| `checkboxGroupInput()` | The state of a group of check boxes (a vector of selected values)                                         |
| `dateInput()`          | A single date                                                                                |
| `dateRangeInput()`     | A range of dates (a vector of the min and max date)                                                      |
| `fileInput()`          | A file                                                                                                   |
| `numericInput()`       | A numeric value                                                                                          |
| `passwordInput()`      | A password                                                                                               |
| `radioButtons()`       | The state of a group of radio buttons (Which option is selected)                                         |
| `selectInput()`        | The current selection(s) for a drop-down list (vector of length 1, or more, that includes only selected values)|
| `sliderInput()`        | A single number/date/date time or a range of the same (a single value, or a vector of the min and max)   |
| `textInput()`          | A character string  (a vector of the entered text)                                                                   |

#### Syntax

`*Input(inputId, label, ...)`

- inputId must be a string, and it must be unique. You cannot use special characters, other than underscores.
  - It creates the id for your input element in HTML and is used to index into the input object in the server
- label describes what the input is for to the user and can use special characters.
- ... means that there are some other specific arguments that are required, depending on the input being used (usually this is something like value or choices).

#### Example

When using a `sidebarLayout()` I typically place most of the inputs into the sidebarPanel and use the mainPanel for my outputs. Notice that the different inputs are separated with commas inside of the sidebarPanel.

```{r eval=FALSE, echo=TRUE}
library(shiny)

ui <- fluidPage(
  sidebarLayout(
    sidebarPanel(actionButton("actionbutton", "Action Button"),
                 checkboxInput("checkboxInput", "Check box", value = TRUE),
                 checkboxGroupInput("checkboxGroupInput", "Check box group", choices = c("A", "B", "C")),
                 dateInput("dateInput", "Date"),
                 dateRangeInput("dateRangeInput", "Date Range"),
                 fileInput("fileInput", "File"),
                 numericInput("numericInput", "Number", value = 10),
                 passwordInput("passwordInput", "Password"),
                 radioButtons("radioButtons", "Radio Buttons", choices = c("A", "B", "C")),
                 selectInput("selectInput", "Drop down", choices = c("A", "B", "C")),
                 sliderInput("sliderInput", "Silder", min = 1, max = 10, value = 5),
                 textInput("textInput", "Text")
                 ),
    mainPanel(
      
      )
    )
  )

server <- function(input, output) { }

shinyApp(ui, server)

```

### Outputs

- Are created with `*Output()` functions.
- An output is composed of two parts: a space on the UI for the output, and code on the server to create the output and render it into HTML. The functions below are used in the UI.
  - We will talk about the server component of output in a moment.
  - You must build the object in the server separately.

| Function             | Inserts              |
|----------------------|----------------------|
| `dataTableOutput()`    | An interactive table |
| `tableOutput()`        | A table              |
| `imageOutput()`        | An image             |
| `plotOutput()`         | A plot               |
| `textOutput()`         | Text                 |
| `verbatimTextOutput()` | Text                 |
| `uiOutput()`           | a Shiny UI element   |
| `htmlOutput()`         | Raw HTML             |

#### Syntax

`*Output(outputId, ...)`

- Output typically only require the outputId.
  - The ID links the output in the UI with the HTML rendered in the server.

#### Example

```{r eval=FALSE, echo=TRUE}
library(shiny)

ui <- fluidPage(
  sidebarLayout(
    sidebarPanel(),
    mainPanel(
      dataTableOutput("dataTableOutput"),
      tableOutput("tableOutput"),
      plotOutput("plotOutput"),
      textOutput("textOutput"),
      verbatimTextOutput("verbatimTextOutput"),
      uiOutput("uiOutput")
      )
    )
  )

server <- function(input, output) { 
  
  output$dataTableOutput <- renderDataTable({ iris })
  
  output$tableOutput <- renderTable({ iris })
  
  output$plotOutput <- renderPlot({ plot(iris) })
  
  output$textOutput <- renderText({ "Some rendered text" })
  
  output$verbatimTextOutput <- renderText({ "Some rendered text" })
  
  output$uiOutput <- renderUI({ selectInput("input", "A rendered UI input", choices = c("A", "B", "C")) })
  
  }

shinyApp(ui, server)

```

- Notice that each output in the UI has a corresponding entry in the server.
- Each output in the server is saved into a `output` object (`output$outputId`), and the UI has a corresponding entry for each ID
- There are different rendering functions used. These turn the outputs into HTML that can be shown in the UI

## Tell the server how to assemble inputs into outputs

### Save object to display to output$outputId

After creating a place in the UI for the output, you need to write the code to create the output. The outputs always need to be saved into the `output` object. This looks like:

```{r eval=FALSE, echo=TRUE}

server <- function(input, output) { 
  
  output$outputId <- # code to create the output, e.g., code that can make a plot, or a table
  
}

```

Every output in the UI needs to have a value created in the server

- for example: if you have `plotOutput("plot")` and `tableOutput("table")` in the UI, you need `output$plot` and `output$table` in the server.

### Build objects with render*()

For an output to be displayed by the webpage, it must be rendered, which means it must be converted into HTML that the webpage can display. Which function you use to render an output depends on its type. 

| Function          | Creates              |
|-------------------|----------------------|
| `renderDataTable()` | An interactive table |
| `renderTable()`     | A table              |
| `renderPlot()`      | A plot               |
| `renderText()`      | A character string   |
| `renderUI()`        | A Shiny UI element   |

```{r eval=FALSE, echo=TRUE}

server <- function(input, output) { 
  
  output$plot <- renderPlot({
    
    plot(iris)
    
  })
  
  output$table <- renderDataTable({
    
    iris
    
  })
  
  # The same table but rendered as a static table. 
  
  # output$table <- renderTable({  
  #   iris
  # })
}

```

### Access inputs with input$inputId

To make outputs interactive, you need to connect them to one or more inputs. The simplest way to do this is to filter the data using the input values. This lets you take static data (like the iris data frame) and transform it into reactive data which can drive outputs.

```{r eval=FALSE, echo=TRUE}
library(shiny)
library(tidyverse)

ui <- fluidPage(
  sidebarLayout(
    sidebarPanel(
      
      # the slider input will return a number between 1 and 100
      sliderInput("n", "Number", min = 1, max = 100, value = 10),
      
      # the select input will return a vector of all the selected species.
      selectInput("species", "Species", 
                  choices = unique(iris$Species),
                  selected = unique(iris$Species),
                  # We are letting the user select multiple values
                  multiple = TRUE)
    ),
    mainPanel(
      tableOutput("table")
    )
    )
  )

server <- function(input, output) { 
  
  output$table <- renderTable({
    
    iris %>% 
      # because species can be a vector of length 1 - 3, we will use the %in%
      # operator to compare
      filter(Species %in% input$species) %>% 
      
      # n will always be a single value
      head(., n = input$n)
    
  })
  
  }

shinyApp(ui, server)

```

## Deploy your app

You can share a shiny app with anyone who has R and the required packages installed on their computer simply by sharing the code, and other required files, with them.

If you want to share a shiny app with the public, or someone who doesn't have R and the required packages, you need to deploy your app to a shiny server. Shinyapps.io provides a free service that lets you upload a limited number of apps (there are paid plans with more active hours that allow for more apps to be loaded). [You can look at this article for details on deploying apps to Shinyapps.io.](https://shiny.rstudio.com/articles/shinyapps.html)


## R&D Personnel App

To run this code, please clone the project using the instructions above. 

```{r eval=FALSE, echo=TRUE}
runApp()
```

You can look at the source code for the shiny app in `CanDev_2022/app.R`. 

- App header
  - Loads required packages.
  - Read in some data from `CanDev_2022/Data` and does some preprocessing

- UI
  - The app uses the layout template, which divides a fluidPage into a sidebar panel, and a main panel.
  - All our inputs are in the sidebar
    - One slider input (`input$year`), one multiple choice select input (`input$category`) and two single choice select inputs (`input$coc`, `input$NAICS`)
  - All the outputs are in the main panel
    - one plot output (`"plot"`) and one table output (`"table"`)
  
- Server
  - In the server we create two entries into the output object using the inputs we created in the UI
    - `output$plot`
      - Using the `renderPlot({ # Code that returns a plot})` function we can turn code that creates a plot into an output that we can display
      - To make the plot interactive, we filter the static data we loaded at the start using the inputs
    - `output$table`
      - Using the `renderDataTable({ # Code that returns a table})` function we can turn code that creates a table (or table like object) into an output that we can display
      - To make the table interactive, we filter the static data we loaded at the start using the inputs.
      
